Modal Abstractions for Virtualizing Memory Addresses

### 2025-07-30 programlogic.tex
slightly refine the why-modalities argument more

\add{
  A program logic for reasoning about code which may work with (and possibly update) multiple address spaces requires dealing with
  several key challenges. It must ensure that reasoning about memory accesses only depends on assertions that hold
  in the active address space at the time of access. It must allow invariants for code or data structures to refer
  to other address spaces. These constraints mean it must also support reasoning about when the active address space
  changes, as this affects which memory assumptions are usable or not for memory access and thus which data structures are
  immediately directly accessible.
}
\add{
  Many approaches which could handle these problems in principle, such as tagging pointers with the relevant address space.
  But such approaches introduce other complexities.
  Most code, even in an operating system kernel, only works with a single address space (the current one).
  Explicitly plumbing address space information through a simple linked list specification, simply because some other part of the program
  may manipulate other address spaces, adds significant specification burden.
  Specifications which need to talk about a particular invariant holding in a specific other address space (or multiple other address spaces) would need to quantify
  over \emph{functions} from address space identity to assertions rather than just assertions.
  Using a modal approach resolves all of these challenges cleanly and uniformly.
  Specifications that are not \emph{about} address space manipulation need not mention an address spaces at all:
  one can use standard separation logic data structure specifications without restructuring,
  and every assertion can be stated for either the current or specific other address spaces.
  In short, modalities make it possible for specifications to mention address spaces when necessary,
  without requiring mention when only one address space is at hand.
}

>>>>

\add{
  A program logic for reasoning about code which may work with (and possibly update) multiple address spaces requires dealing with
  several key challenges. It must ensure that reasoning about memory accesses only depends on assertions that hold
  in the active address space at the time of access. It must allow invariants for code or data structures to refer
  to other address spaces. These constraints mean it must also support reasoning about when the active address space
  changes, as this affects which memory assumptions are usable or not for memory access and thus which data structures are
  immediately directly accessible.
}
\add{
  Many approaches could handle these problems in principle, such as tagging pointers with the relevant address space.
  But such approaches introduce other complexities.
  Most code, even in an operating system kernel, only works with a single address space (the current one).
  Explicitly plumbing address space information through a simple linked list specification, simply because some other part of the program
  may manipulate other address spaces, adds significant specification burden.
  Specifications which need to talk about a particular invariant holding in a specific other address space (or multiple other address spaces) would need to quantify
  over \emph{functions} from address space identity to assertions rather than just assertions.
}

\add{
  Using a modal approach resolves all of these challenges cleanly and uniformly.
  Specifications that are not \emph{about} address space manipulation need not mention address spaces in assertions.
  One can use standard separation logic data structure specifications without restructuring or
  adding explicit address space
  tracking,
  but every assertion can be still stated for either the current or specific other address spaces as needed.
  In short, modalities make it possible for specifications to mention address spaces when
  it is important to the code, and not when it is unimportant to the code.
  \looseness=-1
}
